---
title: "Practical Machine Learning - Course Project"
author: "Akshaya Srivastava"
date: "Sunday, October 26, 2014"
output: html_document
---

Predictive Analytics - Predicting Exercise Qualilty
========================================================

###I. Synopsis
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. Some of the applications of this data range as far as elderly monitoring, life log systems for monitoring energy expenditure, and for supporting weight-loss programs.  

By using data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants, we can attempt to quantify how well dumbbell lifts were performed. With enough data, we will be able to predict the manner in which they did the exercise: exactly according to the specification (Class A), throwing the elbows to the front (Class B), lifting the dumbbell only halfway (Class C), lowering the dumbbell only halfway (Class D) and throwing the hips to the front (Class E)

The *Weight Lifting Exercises Dataset* is harvested from the following paper:

Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. [Qualitative Activity Recognition of Weight Lifting Exercises](http://groupware.les.inf.puc-rio.br/work.jsf?p1=11201). Proceedings of 4th International Conference in Cooperation with SIGCHI (Augmented Human '13) . Stuttgart, Germany: ACM SIGCHI, 2013.


###II. General Housekeeping


```{r}

library(AppliedPredictiveModeling)
library(caret)
library(rattle)
library(rpart.plot)
library(randomForest)

```

###III. Downloading and preprocessing data files

```{r}
# Download Data Files
    download.file("http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv", destfile = "./pml-training.csv")
    download.file("http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv", destfile = "./pml-testing.csv")

# Import data while substituting empty values as 'NA'.
        training <- read.csv("./pml-training.csv", na.strings=c("NA",""), header=TRUE)
        colnames_train <- colnames(training)

        testing <- read.csv("./pml-testing.csv", na.strings=c("NA",""), header=TRUE)
        colnames_test <- colnames(testing)

# Verify that the column names (excluding classe and problem_id) are identical in both the training and the test set.
        all.equal(colnames_train[1:length(colnames_train)-1], colnames_test[1:length(colnames_train)-1])

# Count the number of non-NAs in each column
        nonNAs <- function(x) {
                as.vector(apply(x, 2, function(x) length(which(!is.na(x)))))
                }

# Build vector of missing data or NA columns to drop.
        colcnts <- nonNAs(training)
        drops <- c()
        for (cnt in 1:length(colcnts)) {
                if (colcnts[cnt] < nrow(training)) {
                        drops <- c(drops, colnames_train[cnt])
                        }
                }

# Drop NA data and the first 7 columns as they are unnecessary for prediction.
        training <- training[,!(names(training) %in% drops)]
        training <- training[,8:length(colnames(training))]

        testing <- testing[,!(names(testing) %in% drops)]
        testing <- testing[,8:length(colnames(testing))]

# Remove near zero covariates
        nsv <- nearZeroVar(training,saveMetrics=TRUE)
        head(nsv)

        training <- training[,!nsv$nzv]
        testing <- testing[,!nsv$nzv]
```

Given that all of the near zero variance variables (nsv) are FALSE, there's no need to eliminate any covariates due to lack of variablility.


###IV. Split data to training and testing for cross validation.

```{r}
        inTrain <- createDataPartition(y=training$classe, p=0.7, list=FALSE)
        new_training <- training[inTrain,]
        new_testing <- training[-inTrain,]
```

We got `r dim(new_training)[1]` samples and `r dim(new_training)[2]` variables for training; `r dim(new_testing)[1]` samples and `r dim(new_testing)[2]` variables for testing.


###V. Analysis
####1. Regression Tree

Now we fit a tree to these data, and summarize/plot it. First, we use the 'tree' package because it is much faster than 'caret' package.

```{r}

#install.packages("tree")
library(tree)
set.seed(1990)
        tree.training=tree(classe~.,data=new_training)
        summary(tree.training)

        plot(tree.training)
        text(tree.training, pretty=0, cex =.8)

```

This is a bushy tree, and it needs to be pruned. Print the tree for detailed summary of the tree.

####2. Use Rpart form Caret (very slow processing)


```{r}

library(caret)
set.seed(32323)

        modFit <- train(classe ~ .,method="rpart",data=new_training)
        print(modFit$finalModel)

# Draw a Prettier plot

library(rattle)
fancyRpartPlot(modFit$finalModel)

```

The result from 'caret' 'rpart' package is close to 'tree' package.


####3. Cross Validation

Check the performance of the tree on the testing data by cross validation.

```{r}
        tree.pred=predict(tree.training,new_testing,type="class")
        predMatrix = with(new_testing,table(tree.pred,classe))
        sum(diag(predMatrix))/sum(as.vector(predMatrix)) # error rate for tree
```

The error rate 0.67 is not very accurate.

```{r}
        tree.pred=predict(modFit,new_testing)
        predMatrix = with(new_testing,table(tree.pred,classe))
        sum(diag(predMatrix))/sum(as.vector(predMatrix)) # error rate for caret
```

The error rate 0.50 from 'caret' package is much lower than the result from 'tree' package.


####4. Pruning the tree

Use Cross Validation to prune the tree.

```{r}
        cv.training=cv.tree(tree.training,FUN=prune.misclass)
        cv.training

        plot(cv.training)
```


The plot shows that when the size of the tree goes down, the deviance goes up. It means the 16 is a good size (i.e. number of terminal nodes) for this tree. We do not need to prune it.

Suppose we prune it at size of nodes at 12.

```{r}
        prune.training=prune.misclass(tree.training,best=12)
```


Now lets evaluate this pruned tree on the test data.


```{r}
        tree.pred=predict(prune.training,new_testing,type="class")
        predMatrix = with(new_testing,table(tree.pred,classe))
        sum(diag(predMatrix))/sum(as.vector(predMatrix)) # error rate
```


The error rate 0.62 is a little less than 0.67, so pruning did not hurt us with respect to misclassification errors, and gave us a simpler tree. We use less predictors to get almost the same result. By pruning, we got a shallower tree, which is easier to interpret.

The single tree is not good enough, so we should use random forests.


###VI. Random Forests

These methods use trees as building blocks to build more complex models.

####1. Random Forests

Random forests build lots of bushy trees and then average them to reduce the variance.


```{r}

library(randomForest)
set.seed(1990)

#Lets fit a random forest and see how well it performs.

        rf.training=randomForest(classe~.,data=new_training,ntree=100, importance=TRUE)
        rf.training

        #plot(rf.training, log="y")
        varImpPlot(rf.training,)

        #rf.training1=randomForest(classe~., data=new_training, proximity=TRUE )
        #DSplot(rf.training1, new_training$classe)

```

we can see which variables have higher impact on the prediction.


###VII. Out-of Sample Accuracy

Our Random Forest model shows OOB (out-of-bag) estimate of error rate: 0.61% for the training data. Now we will predict it for out-of sample accuracy.

Lets evaluate this tree on the test data.

```{r}

        tree.pred=predict(rf.training,new_testing,type="class")
        predMatrix = with(new_testing,table(tree.pred,classe))
        sum(diag(predMatrix))/sum(as.vector(predMatrix)) # error rate

```


*0.995582 means we got a very accurate estimate.*

No. of variables tried at each split: 7. It means every time we only randomly use 7 predictors to grow the tree. Since p = 53, we can have it from 1 to 53, but it seems 7 is enough to get the good result.



###VIII. Conclusion

Now we can predict the testing data from the website.

```{r}

        answers <- predict(rf.training, testing)
        print(answers)

```

It shows that this random forest model performed a better job.


.